Paradigms in Physics 2.0

### 2017-06-28 paper.tex
before changes with feedback from meeting

\newcommand\corinnesays[1]{\textcolor{red}{[\sc CAM: \em{#1}]}}

@@

\title{Paradigms in Physics 2.0}
\author{David Roundy}
\author{Liz Gire}
\author{Ethan Minot}
\author{Emily van Zee}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}
\author{Corinne A. Manogue}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}

@@

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  This paper will describe both the process by which our
department achieved consensus on this curricular change, and the
resulting curriculum.  The Paradigms 2.0 committee begain with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include elimination or major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students to for their
upper-division courses.
\end{abstract}

@@

\paragraph{Transfer students}
Our curriculum is strongly affected by our desire to accomodate
transfer students from community colleges.  This leads us to focus on
a physics major in which all upper-division courses are taken in two
years.  However, our current curriculum is very hard on students in
the fall of their junior year, and especially so for transfer
students.  So we aimed to structure and order the courses such that non-transfer
students could be better prepared for their junior year, while
transfer students could postpone a few courses so that their junior
Fall would be no more difficult than that of any other student.

\paragraph{Coupled curricular changes}
A backlog of curricular changes have accumulated that could not be
separated from an examination of the curriculum as a whole.  Most
notable were discussions of the requirements for computational
physics, the amount of required electronics credits, and the role of
our modern physics course.  These persistent issues were not easily
addressed without addressing the physics major curriculum as a whole,
and any discussion of one tended to devolve into a discussion of
everything, without progress being made.  
\corinnesays{This paragraph could perhaps be tightened.}

@@

% David R, Ethan, Oksana, Bo, Guenter, Weihong, Matt, Yun-Shik, Heidi,
% Liz, Davide = 11
Only 5 of our 16 current tenure-line faculty were involved in the
orignal Paradigms effort.  Some of the newest faculty have only a
superficial undertanding of our course structure.  This poses a
challenge when these professors teach courses that are closely
intertwined with one another.

\section{Paradigms 2.0 process}
We began the Paradigms 2.0 process in the Winter of 2016, having in
the previous year obtained faculty agreement, and support from our
Department head.  The process was spearheaded by a committee of four
(the authors DR, LG, EM, and CAM, with EvZ present at meetings
documenting our process).  During the Winter quarter, we informed the
community of our process through a colloquium, and collected student
and faculty perspectives on the existing curriculum.  We then
interviewed faculty who recently taught each of our existing courses
to document topics that were currently covered.  This resulted in a
total of $\sim$700 index cards in $\sim$30 stacks, with each stack
corresponding to one course, and each card describing a topic,
color-coded as in Fig.~\ref{fig:schedule}.

@@

\subsection{Math bits}
We chose to eliminate the Math Methods course in favor of formal
``math bits'' in the Paradigms.  Our previous Math Methods Capstone
posed a challenge regarding placement in the curriculum: it needed to
be sufficiently late for students to be adequately prepared
for advanced mathematical content, but in order to be functional in
the curriculum it should teach content that is required by other
physics courses, and thus should be earlier in the curriculum.  In
response to these issues, and in recognition that the majority of our
majors are not bound for graduate school, and have no need for
advanced math methods, we chose to integrate the needed math content
into the Paradigms, and allow advanced students to take our
graduate-level Math Methods course.  In order to ensure that our
students to receive the math instruction they require, and based on
our previous experience with math-intensive weeks (the ``odd'' weeks)
in the Paradigms, we chose to integrate one week of ``math bits'' into
each Paradigm.  The math bits for the entire year are treated as a
distinct teaching assignment, so students will be accustomed to the
treatment of math as a distinct subject supporting the physics
content.  This provides some relief from teaching for the Paradigms
professors, and at the same time ensures that they do not succumb to
the temptation to short-change the math content to the detriment of
the students.
\corinnesays{This paragraph is awkward.  I'm still thinking about some rewording.  This might be a good paragraph to discuss this afternoon.}

@@

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Challenges in
  Contemporary Physics}, and \emph{Theoretical Mechanics}.  Both of
these courses ramp up student mathematical abilities prior to their
junior year.  The Challenges course focuses on estimation,
dimensional reasoning, and interpretation of integrals, while
Theoretical Mechanics will teach power series approximations, and
expose students to increased levels of mathematical sophistication.

These courses \emph{will} have the challenge (which was removed from
the computational course!) of teaching both juniors and sophomores
together.  They will be taught in the winter and spring quarters, so
as to reduce the burden on transfer students in the Fall.  They were
explicitly constructed to \emph{not} teach any content required for
Fall or Winter Paradigms so that they can be taken concurrently with
the Paradigms.

@@

While contemporary challenges determine the narrative flow of the
course, physics concepts and physical reasoning skills are the main
substance. Students are introduced to thermodynamics, statistical
mechanics, electromagnetic radiation, quantum mechanics, and modern
experimental physics---in each case motivated by one or more
``contemporary challenges''. Together with new physics concepts,
students are given new reasoning tools, such as the equipartition
theorem, the quantum-classical correspondence principle, order of
magnitude calculations, simplifying assumptions, and numerical
integration. Mathematically rigorous derivations are only briefly
discussed in class, with additional details offered in online video
format. In summary, we emphasis the physics concepts and reasoning
skills that allow professional physicists to quickly/quantitatively
make an initial assessment of a complex problem.

The students engage with the material by solving daily quiz questions
in class, solving weekly homework problems, performing a modern
physics experiment, and writing a term paper. The daily quiz questions
help clarify the goal of each lecture. The weekly homework questions
are directly related to contemporary challenges are require a variety
of estimation techniques. The modern physics experiment, performed
with a low cost kit (LEDs, diffraction grating, rulers, volt meter
etc.) emphasizes a quantitative approach to observation and explores
issues of experimental uncertainty. Lastly, the term paper allows
students to dive deeper into a particular contemporary challenge. The
term paper tests student's ability to make their own initial
assessment of a complex problem, and communicate their ideas with
diagrams, graphs and estimations.

@@

While the Challenges courses takes a more experimental/applied physics
focus, the Methods of Theoretical Mechanics course has a more
theoretical physics flavor. The theme of this course is the discussion
of strategies for making sense of physics problems and symbolic
problem solving. This sense-making includes coordinating and
interpreting symbolic expressions with conceptual understandings,
geometric relationships, and physical intuitions.

This Theoretical Mechanics course is aimed at students who are taking
or have completed the last course in the introductory sequence. The
physics content of the course is advanced Newtonian mechanics,
introduction to Lagrangian and Hamiltonian techniques, and special
relativity. These topics at this stage for the students are convenient
for explicit discussion of sense-making strategies in physics because
(a) students are have recently studied problems that serve as limiting
cases for more complex problems, (b) students at this level are
transitioning from solving problems primarily involving numbers to
problems with only symbolic parameters, and (c) we can discuss
sense-making as strategies for approaching a problem, evaluating
answers, and refining intuitions about relativistic and other
unfamiliar situations. For example, we teach students how to use
spacetime diagrams to develop a story about a relativistic situation
and how to use hyperbola trig on these diagrams to perform Lorentz
transformations that can be checked against the results of algebraic
Lorentz transformations.  The sense-making goals of the course are on
equal footing with the physics goals, and sense-making is integrated
explicitly in all aspects of the course: exams, homework and during
in-class activities. Our aim is that students will develop
sense-making skills that will improve their learning in advanced
courses and will impress their upper-division course instructors,
research advisors, and future employers.

@@

\section{Summary}
We have developed a significant incremental change to the Paradigms
curriculum.  This change focused on the first two big ideas underlying
the Paradigms: (1) close attention to content ordering, and (2)
developing a consensus and understanding of the curriculum among our
faculty members.  Our commitment to using active engagement in our
upper-division curriculum remains unchanged.  While the specific
content ordering we developed is interesting, the \emph{process} by
which we reached it, and developed a faculty consensus is at the
essence of what makes distinctive the Paradigms in Physics program.
The process of developing faculty understanding of the Paradigms is
ongoing, as we are having professors new to the Paradigms shadow
experienced faculty as they teach in the Paradigms, prior to teaching
the courses themselves..

>>>>

\newcommand\corinnesays[1]{\textcolor{red}{[\sc CAM: \em{#1}]}}
\newcommand\davidsays[1]{\textcolor{blue}{[\sc DR: \em{#1}]}}

@@

\title{Paradigms in Physics 2.0}
\author{David Roundy}
\author{Elizabeth Gire}
\author{Ethan Minot}
\author{Emily van Zee}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}
\author{Corinne A. Manogue}
\affiliation{Department of Physics, Oregon State University, Corvallis, Oregon, 97331}

@@

\begin{abstract}
In 2016, the Department of Physics at Oregon State University began a
process to revise our Paradigms in Physics curriculum for physics
majors.  This paper will describe both the process by which our
department achieved consensus on this curricular change, and the
resulting curriculum.  The Paradigms 2.0 committee begain with a
survey of students and faculty, followed by individual interviews with
the faculty teaching each existing course.  As we developed a plan to
address student- and faculty-identified challenges in the curriculum,
we met with each faculty member individually to explain and refine our
proposal, which was unanimously approved by the faculty.  Major
changes include major changes to several courses (math
methods, computational physics, modern physics, electronics, and
classical mechanics), including the introduction of two sophomore-year
courses designed specifically to help prepare students to for their
upper-division courses.
\end{abstract}

@@

\paragraph{Transfer students}
Our curriculum is strongly affected by our desire to accommodate
transfer students from community colleges.  This leads us to focus on
a physics major in which all upper-division courses are taken in two
years.  However, our current curriculum is very hard on students in
the fall of their junior year, and especially so for transfer
students.  So we aimed to structure and order the courses such that non-transfer
students could be better prepared for their junior year, while
transfer students could postpone a few courses so that their junior
Fall would be no more difficult than that of any other student.

\paragraph{Coupled curricular changes}
A backlog of curricular changes have accumulated that could not be
separated from an examination of the curriculum as a whole.  Most
notable were discussions of the requirements for computational
physics, the amount of required electronics credits, and the role of
our modern physics course.  These persistent issues were not easily
addressed without addressing the physics major curriculum as a whole,
everything, without progress being made.  
\corinnesays{This paragraph could perhaps be tightened.}

@@

% David R, Ethan, Oksana, Bo, Guenter, Weihong, Matt, Yun-Shik, Heidi,
% Liz, Davide = 11
Only 5 of our 16 current tenure-line faculty were involved in the
orignal Paradigms effort.  Some of the newest faculty have only a
superficial understanding of our course structure.  This poses a
challenge when these professors teach courses that are closely
intertwined with one another.

\section{Paradigms 2.0 process}
We began the Paradigms 2.0 process in the Winter of 2016, having in
the previous year obtained faculty agreement, and support from our
Department head.  The process was spearheaded by a committee of four
(the authors DR, EG, EM, and CAM, with EvZ present at meetings
documenting our process).  During the Winter quarter, we informed the
community of our process through a colloquium, and collected student
and faculty perspectives on the existing curriculum.  We then
interviewed faculty who recently taught each of our existing courses
to document topics that were currently covered.  This resulted in a
total of $\sim$700 index cards in $\sim$30 stacks, with each stack
corresponding to one course, and each card describing a topic,
color-coded as in Fig.~\ref{fig:schedule}.

@@

\subsection{Math bits}
We chose to eliminate the Math Methods course in favor of formal
``math bits'' in the Paradigms.  Our previous Math Methods Capstone
posed a challenge regarding placement in the curriculum: it needed to
be sufficiently late for students to be adequately prepared
for advanced mathematical content, but in order to be functional in
the curriculum it should teach content that is required by other
physics courses, and thus should be earlier in the curriculum.  In
response to these issues, and in recognition that the majority of our
majors are not bound for graduate school, and have no need for
advanced math methods, we chose to integrate the needed math content
into the Paradigms, and allow advanced students to take our
graduate-level Math Methods course.  In order to ensure that our
students to receive the math instruction they require, and based on
our previous experience with math-intensive weeks (the \davidsays{``extra''
weeks having 3-week courses fitted into a 10-week quarter})
in the Paradigms, we chose to integrate one week of ``math bits'' into
each Paradigm.  The math bits for the entire year are treated as a
distinct teaching assignment, so students will be accustomed to the
treatment of math as a distinct subject supporting the physics
content.  This provides some relief from teaching for the Paradigms
professors, and at the same time ensures that they do not succumb to
the temptation to short-change the math content to the detriment of
the students.
\corinnesays{This paragraph is awkward.  I'm still thinking about some rewording.  This might be a good paragraph to discuss this afternoon.}

@@

\subsection{Sophomore courses}
We introduced two new sophomore-level courses: \emph{Physics of
Contemporary Challenges} and \emph{Techniques of Theoretical Mechanics}.  Both of
these courses ramp up student mathematical abilities prior to their
junior year.  The Challenges course focuses on estimation,
dimensional reasoning, and interpretation of integrals, while
Theoretical Mechanics teaches power series approximations and
exposes students to increased levels of mathematical sophistication.

These courses have the challenge (which was removed from
the computational course!)\corinnesays{remove this parenthetical comment} 
of teaching both juniors and sophomores
together.  They are taught in the winter and spring quarters, so
as to reduce the burden on transfer students in the Fall.  They were
explicitly constructed to \emph{not} teach any content required for
Fall or Winter Paradigms so that they can be taken concurrently with
the Paradigms.

@@

While contemporary challenges determine the narrative flow of the
course, physics concepts and physical reasoning skills are the main
substance. Students are introduced to thermodynamics, statistical
mechanics, electromagnetic radiation, quantum mechanics, and modern
experimental physics---in each case motivated by one or more
``contemporary challenges''. Together with new physics concepts,
students are given new reasoning tools, such as the equipartition
theorem, the quantum-classical correspondence principle, order of
magnitude calculations, simplifying assumptions, and numerical
integration. Mathematically rigorous derivations are only briefly
discussed in class, with additional details offered in online video
format. In summary, we emphasize the physics concepts and reasoning
skills that allow professional physicists to quickly/quantitatively
make an initial assessment of a complex problem.

The students engage with the material by solving daily quiz questions
in class, solving weekly homework problems, performing a modern
physics experiment, and writing a term paper. The daily quiz questions
help clarify the goal of each lecture. The weekly homework questions
are directly related to contemporary challenges requiring a variety
of estimation techniques. The modern physics experiment, performed
with a low cost kit (LEDs, diffraction grating, rulers, volt meter
etc.) emphasizes a quantitative approach to observation and explores
issues of experimental uncertainty. Lastly, the term paper allows
students to dive deeper into a particular contemporary challenge. The
term paper tests student's ability to make their own initial
assessment of a complex problem, and communicate their ideas with
diagrams, graphs and estimations.

@@

While the Challenges course takes a more experimental/applied physics
focus, the Methods of Theoretical Mechanics course has a more
theoretical physics flavor. The theme of this course is the discussion
of strategies for making sense of physics problems and symbolic
problem solving. This sense-making includes coordinating and
interpreting symbolic expressions with conceptual understandings,
geometric relationships, and physical intuitions.

The Theoretical Mechanics course is aimed at students who are taking
or have completed the last course in the introductory sequence. The
physics content of the course is advanced Newtonian mechanics,
introduction to Lagrangian and Hamiltonian techniques, and special
relativity. These topics at this stage for the students are convenient
for explicit discussion of sense-making strategies in physics because
(a) students are have recently studied problems that serve as limiting
cases for these more complex problems, (b) students at this level are
transitioning from solving problems primarily involving numbers to
problems with only symbolic parameters, and (c) sense-making as strategies 
can be discussed for approaching a problem, evaluating
answers, and refining intuitions about relativistic and other
unfamiliar situations. For example, we teach students how to use
spacetime diagrams to develop a story about a relativistic situation
and how to use hyperbola trigonometry on these diagrams to perform Lorentz
transformations that can be checked against the results of algebraic
Lorentz transformations.  The sense-making goals of the course are on
an equal footing with the physics goals, and sense-making is integrated
explicitly in all aspects of the course: exams, homework and during
in-class activities. Our aim is that students will develop
sense-making skills that will improve their learning in advanced
courses and will be valued by their upper-division course instructors,
research advisors, and future employers.

@@

\section{Summary}
We have developed a significant change to the Paradigms
curriculum.  This change focuses on the first two big ideas underlying
the Paradigms: (1) close attention to content ordering, and (2)
developing a consensus and understanding of the curriculum among our
faculty members.  Our commitment to using active engagement in our
upper-division curriculum remains unchanged.  While the specific
content ordering we developed may be interesting, the \emph{process} by
which we reached it and developed a faculty consensus is at the
essence of what makes the Paradigms in Physics program distinctive.
Emily van Zee, a researcher with experience in ..., 
\corinnesays{Emily, please add a short description of yourself} 
has completed a retrospective study of the process we used to develop the 
original Paradigms program and we have used her results extensively in 
designing and implementing the Paradigms 2.0 process.  She has continued 
to observe, study, and document our work as we have begun implementing 
the revised curriculum. 
\corinnesays{David:  Add a citation to http://physics.oregonstate.edu/portfolioswiki/toolkit:start}
The process of developing faculty understanding of the Paradigms is
ongoing, as we are having professors new to the Paradigms shadow
experienced faculty as they teach in the Paradigms, prior to teaching
the courses themselves.
